ODGI: scalable tools for pangenome graphs

### 2021-06-01 manuscript/main.tex
Shortened

\newcommand{\odgi}{ODGI}
\newcommand{\FIXME}{\textbf{!!FIXME!!}}
\newcommand{\cmd}[1]{{\scriptsize\textrm{#1}}}
\newcommand{\cmdbf}[1]{{\textbf{#1}}}
\newcommand{\topic}[1]{{\cmdbf{#1}}:}
\newcommand{\Rplus}{\protect\hspace{-.1em}\protect\raisebox{.35ex}{\smaller{\smaller\textbf{+}}}}
\newcommand{\pp}{\Rplus\Rplus}
\newcommand{\Cpp}{\mbox{C\Rplus\Rplus}\xspace}

@@

    A pangenome models the full set of genomic elements in a population~\citep{32453966}. Indeed, these data structures
    encode the mutual relationships between all the genomes represented, in contrast to reference-based approaches
    which relate sequences only to a particular genome, chosen as reference. A class of methods to represent
    pangenomes involves the \textit{sequence graphs}~\citep{2488477}, where similar orthologous regions between genomes
    collapse into a single representative sequence. In \textit{node-labeled} sequence graphs, nodes indicate DNA
    sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph. A
    \textit{bidirected} sequence graph represents both strands of DNA. On this model, \textit{variation graphs} add the
    concept of \textit{path} to embed linear sequences (e.g., genomes or haplotypes) into the graph~\citep{30125266}.
    Paths provide a stable coordinate system, allowing graph annotations and interoperability between different graphs.

    Thanks to advances in sequencing technologies, new genome assemblies are produced at a high rate. This amount of
    data offers the opportunity to study genomic variation as never before, but it also brings challenges in how to
    represent and manipulate hundreds of genomes at a gigabase scale. Variation graphs compactly represent the full
    genetic variation across a population, but graph-based data structures present computational overheads: in addition
    to the sequences, they must also represent the graph topology. Several genome data structures were developed to
    address these challenges~\citep{33040146}. Memory-efficient representations do not allow dynamic updates to the
    graph. Furthermore, the changes must perform efficiently on the complex graphs generated from whole-genome
    alignment of vertebrates. Highly repetitive regions, such as centromeres, segmental duplications, and acrocentric
    chromosomes lead to graphs which present very high-depth nodes. The node depth is defined as the
    number of times in which the node is crossed by all the paths embedded in the graph. High depth increases the
    complexity of the operations performed on the paths and the nodes they cross.

    To overcome this limitation, we have developed an Optimized Dynamic Genome/Graph Implementation (ODGI). Each tool
    loads graphs into the succinct dynamic HandleGraph model~\citep{33040146} and applies methods based on the
    HandleGraph API. Several algorithms are shared with the VG toolkit (VG)~\citep{30125266}, the state-of-the-art
    toolkit for working with pangenome graphs. However, although similar in scope to VG, \odgi\ importantly differs in
    its design to support highly complex pangenome graphs. The majority of the \odgi\ tools are implemented in an
    index-free manner, avoiding the burning task to create an index structure. Indeed, index-constrained
    implementations require building indexes also for single or little operations on the graphs. Moreover,
    thanks to its efficient path representation, \odgi\ tools can build the internal graph representation (the \odgi\
    format) in parallel, which is one of the major bottlenecks when working with very large and deep graphs,
    scaling up to hundreds of large vertebrate genomes.

    \odgi\ is entirely focused on "first-order" operations on the graph
    (Table~\ref{tab:01}), providing, unlike VG, no method to construct
    graphs from raw sequences, map reads, or call variants. It
    supports interrogation of the graph for basic dimensional
    properties, positional and path-related queries, and visualisation
    in one-dimension (1D) and two-dimensions (2D). \odgi\ also supports
    operations on the graph such as subset, subdivide, break, combine,
    normalize, or order its components, nodes, and paths. Most of the
    tools are designed to be applied together, piping the output of
    one tool into the next, thereby preventing the creation of
    intermediate files, and reducing the number of IO operations.

    \section{Methods}

    \odgi\ is designed to build and modify paths in parallel, keeping its representation in memory efficient. It
    applies concepts first introduced in the dynamic version of the Graph BWT~\citep{31406990}, a scalable
    implementation of the graph extension of the positional Burrows-Wheeler transform~\citep{28702075}. Specifically,
    we build a \textit{node-local} model of the graph, where the graph is stored in a vector of node structures.
    The identifier of each node is its offset in this vector plus 1. Each structure contains the sequence of the
    node, its edges in both directions (forward and backward), and a vector of \textit{path steps} that describes
    how the embedded paths walk across the node. Each step records the path identifier and the previous and next
    path steps on the same path. To reduce the encoding, we encode path steps using a local alphabet that maps the
    \textit{N} neighbors of the node into the range \textit{1..N}. To further save space, \textit{node deltas},
    rather than IDs, are stored in this alphabet. The delta between two nodes is defined as their distance in the
    graph vector (i.e., the difference between the node offsets). The path step vector, of length \textit{n}, is
    stored using a dynamic succinct integer vector that requires \textit{O(5nw)} bits, where \textit{w} is
    \textit{\textasciitilde log\textsubscript{2}(N)}. Finally, to allow operating on path steps in parallel, each node
    structure includes a byte-width mutex lock.

    Table~\ref{tab:02} shows a comparison of the operations necessary to extract a subgraph using \odgi\ and VG.

    %TODO to repeat on a bigger machine to avoid vg crashing
    \begin{table}[!t]
      \processtable{Example of performance improvement between
        \cmd{odgi} and \cmd{VG} equivalent commands with output file
        sizes. The centromere region was extracted from CHR 6 of the
        HPRC year one assembly\citep{HPRC}, i.e., 88 haploid, phased
        human genome assemblies from 44 individuals plus the chm13
        cell line and GRCh38 reference genomes stored in 4.3GB GFA
        format\citep{GFA}.
        %
        First GFS was converted from GFA to \odgi\ and \cmd{vg}
        formats, respectively.  \odgi's path implementation can
        process paths in parallel, outperforming \cmd{vg}. For the
        same reason \cmd{odgi extract} is $20\times$ faster than
        \cmd{vg chunk} extracting a portion from the full
        graph. \FIXME: vg never finished.
        %
        All timings were performed on a server class machine
        on SSD MODEL with X GB of RAM and 48 CPU cores (Y x 12 core
        AMD Opteron) Processor MODEL @ 3.3 GHz with 8MB L2
        Cache.\FIXME
          \label{tab:02}} {
          \resizebox{.99\columnwidth}{!}{
            \begin{tabular}{@{}lllll@{}}
                \toprule Operation & Tool & Runtime (mm:ss) & Memory (GB) & Output size (GB)    \\
                \midrule
                Format conversion                           &       &       &                   \\
                                    & \cmd{odgi build}      & 1:35  & 10.39 & 5.4               \\
                                    & \cmd{vg convert}      & 8:14  & 28.43 & 6.1               \\
                Subgraph extraction                         &       &       &                   \\
                                    & \cmd{odgi extract}    & 1:15  & 9.43  & 2.7               \\
                                    & \cmd{vg chunk}        & 21:09 & 59.33 & 1.7               \\
                \botrule
            \end{tabular}}}

    \section{Features}

    \odgi\ has a wide range of commands.

@@

    \section{Discussion}
    We implemented a tool suite, \odgi\, which is simple to work with to analyse and manipulate pangenome graphs at a
    population scale. ODGI's path implementation allows working with graphs of any complexity, without particular
    limitations. We have demonstrated that \odgi\ tools efficiently perform with pangenome graphs embedding full human
    chromosomes. \odgi\ is not only a tool suite but also a library, already used in pangenome graphs building
    (\citep{pggb}) and normalization (\citep{smoothxg}). Furthermore, users can add new tools, exploiting ODGI as a
    framework where to implement algorithms working on variation graphs. Therefore, \odgi\ offers an efficient and
    extendable suite of tools ready for the upcoming challenges in the analyses of hundreds of genomes at a gigabase
    scale that characterize the pangenomic era.

@@

    \textit{Conflict of Interest}: none declared.
    \vspace*{+12pt}

>>>>

\newcommand{\odgi}{ODGI}
\newcommand{\vg}{VG}

\newcommand{\FIXME}{\textbf{!!FIXME!!}}
\newcommand{\cmd}[1]{{\scriptsize\textrm{#1}}}
\newcommand{\cmdbf}[1]{{\textbf{#1}}}
\newcommand{\topic}[1]{{\cmdbf{#1}}:}
\newcommand{\Rplus}{\protect\hspace{-.1em}\protect\raisebox{.35ex}{\smaller{\smaller\textbf{+}}}}
\newcommand{\pp}{\Rplus\Rplus}
\newcommand{\Cpp}{\mbox{C\Rplus\Rplus}\xspace}

@@

    Thanks to advances in sequencing technologies, new genome
    assemblies are produced at a high rate, including the human
    pangenome reference consortium (HPRC) and telomere-to-telemere
    efforts, e.g.~\citep{Miga:2020}.  A pangenome models the full set
    of genomic elements in a given species or clade. Pangenomics thus
    stands in contrast to reference-based genomic approaches which
    relate sequences to a particular consensus model of the
    genome~\citep{Eizenga:2020}. A pangenome can elegantly be
    represented by a graph data structure incorporating sequences as
    nodes and their connections as edges. These nodes can be shared
    for homologs and orthologs.  A `bidirected' graph represents both
    strands of DNA and so-called paths represent contigs that run
    through the nodes. A path can represent a chromosome or part
    thereof.

    A major challenge is writing software that can deal with the size
    of these graphs. vgtools pioneered graph
    processing~\citep{vgtools,Eizenga:2020b}, but was not designed for
    processing large data (see Table~\ref{tab:02}). With \odgi\ we
    wrote a new toolset and API in \Cpp\ named Optimized Dynamic
    Genome/Graph Implementation (ODGI).  that implements is a memory
    efficient graph structure that can be dynamically updated using
    multiple CPUs. This is particularly relevant for highly repetitive
    regions. Segmental duplications, and acrocentric chromosomes lead
    to graphs which present very high node depths which greatly
    increases processing complexity.

    Note that \odgi\ is focused on `first-order' graph
    operations.  Other tools are required to construct graphs from raw
    sequences, map reads, or call variants.  \odgi\ provides
    operations on an existing graph, such as subset, subdivide, break,
    combine, normalize and order its components, nodes, and
    paths. Most of the tools are designed to be applied together,
    piping the output of one tool into the next.

    \vspace{-0.1in}

    \section{Results}

    \odgi\ is designed to build and modify paths in parallel. It
    applies concepts first introduced in the dynamic version of the
    graph extension of the positional Burrows-Wheeler
    transform~\citep{Siren:2020}. For implementation details see the
    \odgi\ online documentation. \odgi\ includes the following tools

@@

    \section{Discussion}

    Pangenome graphs stand to become a ubiquitous tool in
    genomics\citep{Eizenga:2020}.  With \odgi\ we implemented a
    state-of-the-art tool suite, that can transform, analyse and
    visualise pangenome graphs at scale.  \odgi\ also exposes an API
    that is used by other tools for pangenome graph building and
    normalization. Future work will add support for RNA and protein
    sequences and expand on metadata capabilities of large pangenome
    graphs.

    %TODO to repeat on a bigger machine to avoid vg crashing
    \begin{table}[th]
      \processtable{\small
        Example of performance improvement between
        \cmd{odgi} and \cmd{vg}~\citep{vgtools} equivalent commands with output file
        sizes. The centromere region was extracted from CHR 6 of the
        HPRC year one assembly, i.e., 88~haploid, phased
        human genome assemblies from 44~individuals plus the chm13
        cell line and GRCh38 reference genomes stored in 4.3GB~GFA
        format\citep{GFA}.
        %
        First GFS was converted from GFA to \odgi\ and \vg\
        formats, respectively.  \odgi's path implementation can
        process paths in parallel, outperforming \vg. For the
        same reason \cmd{odgi extract} is $20\times$ faster than
        \cmd{vg chunk} extracting a portion from the full
        graph. \FIXME: vg never finished.
        %
        All timings were performed on a server class machine
        on SSD MODEL with X GB of RAM and 48 CPU cores (Y x 12 core
        AMD Opteron) Processor MODEL @ 3.3 GHz with 8MB L2
        Cache.\FIXME
        % \vspace{0.5in}
        \label{tab:02}}
                   {
                     % \resizebox{.8\textwidth}{!}{
            \begin{tabular*}{.5\textwidth}{@{}lllll@{}}
                \toprule Operation & Tool & Runtime (mm:ss) & Memory (GB) & Output size (GB)    \\
                \midrule
                Format conversion                           &       &       &                   \\
                                    & \cmd{odgi build}      & 1:35  & 10.39 & 5.4               \\
                                    & \cmd{vg convert}      & 8:14  & 28.43 & 6.1               \\
                Subgraph extraction                         &       &       &                   \\
                                    & \cmd{odgi extract}    & 1:15  & 9.43  & 2.7               \\
                                    & \cmd{vg chunk}        & 21:09 & 59.33 & 1.7               \\
                \botrule
            \end{tabular*}} %}

@@

    \textit{Conflict of Interest}: none declared.
    % \vspace*{+12pt}
